Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-05-01 ms_pre_post.Rmd
Add Matthew's comments

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "28 April 2017"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Inbreeding depression is widely hypothesised to drive adaptive evolution of pre-copulatory and post-copulatory mechanisms of inbreeding avoidance, which in turn are hypothesised to affect evolution of polyandry (i.e., female multiple mating). However, surprisingly little theory or modelling critically examines selection for pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, given evolutionary constraints and direct costs, or examines how evolution of inbreeding avoidance strategies might feed back to affect evolution of polyandry. Selection for post-copulatory inbreeding avoidance, but not for pre-copulatory inbreeding avoidance, requires polyandry, while interactions between pre-copulatory and post-copulatory inbreeding avoidance might cause functional redundancy (i.e., 'degeneracy') potentially generating complex evolutionary dynamics among inbreeding strategies and polyandry. We use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression and different evolutionary constraints and direct costs. We find that evolution of post-copulatory inbreeding avoidance causes selection for increased polyandry, and that selection for a costly inbreeding avoidance strategy was weak or negligible given a cost free alternative strategy. Further, fixed pre-copulatory inbreeding avoidance generally precluded evolution of polyandry and therefore post-copulatory inbreeding avoidance, but fixed post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance. Evolution of inbreeding avoidance phenotypes and associated polyandry are likely to be affected by evolutionary feedbacks and degeneracy. All else being equal, while evolution of pre-copulatory inbreeding avoidance and associated low polyandry is more likely when pre-copulatory inbreeding avoidance is precluded or costly, evolution of post-copulatory inbreeding avoidance greatly facilitates evolution of costly polyandry.

@@

From the perspective of a breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), or by biasing fertilisation towards unrelated males after mating has occurred (i.e., post-copulatory inbreeding avoidance). Evolution of post-copulatory inbreeding avoidance requires that females express some degree of polyandry, defined as mating with multiple males during a single reproductive bout. Further, given inbreeding depression, such polyandry might itself be under selection and evolve if it allows females to mate with additional unrelated males following an initial mating with a relative. Polyandry can thereby facilitate pre-copulatory inbreeding avoidance even without any post-copulatory female choice or otherwise biased fertilisation among sperm provided by females' mates (i.e., 'fair raffle'). Overall, polyandry can therefore simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. Resulting indirect selection on polyandry to increase offspring fitness could explain its evolution in cases where multiple mating does not increase female reproductive success, or even reduces reproductive success and therefore imposes direct fitness costs on polyandrous females. However, despite such widely-invoked hypotheses, there is surprising little theory or modelling that critically examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or that examines how the evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry.

Despite the paucity of theory, numerous empirical studies on diverse species have tested for, and in some cases found evidence of, female inbreeding avoidance in systems where polyandry also occurs [@Tregenza2002; @Varian-Ramos2012; @Kingma2013; @Arct2015; but see @Reid2015a]. However, relatively few studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Pre-copulatory inbreeding avoidance has been reported in diverse species, spanning sweet potato weevils [*Cylas formicarius*, @Kuriwada2011], purple-crowned fairy-wrens [*Malurus coronatus*, @Kuriwada2011], and a tropical butterfly [*Bicyclus anynana*, @Fischer2015], while post-copulatory inbreeding avoidance has been reported in red junglefowl [*Gallus gallus*, @Pizzari2004] and crickets [*Teleogryllus oceanicus*, @Simmons2006; *Gryllus bimaculatus*, @Bretman2009]. Evidence for both pre-copulatory and post-copulatory inbreeding avoidance is available across different studies of Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. Meanwhile, @Liu2014 found evidence of pre-copulatory but not post-copulatory inbreeding avoidance within a single study of cabbage beetles (*Colaphellus bowringi*). However, @Ala-Honkola2011 and @Tan2012 found no evidence for pre-copulatory or post-copulatory inbreeding avoidance in fruit flies (*Drosophila melanogaster*), respectively, and Reid et al. [-@Reid2014; -@Reid2015] found no net inbreeding avoidance in song sparrows (*Melospiza melodia*) despite strong inbreeding depression. Taken together, these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature alongside polyandry. However, there is as yet no theory that predicts what combinations of pre-copulatory and post-copulatory inbreeding avoidance and polyandry should be favoured by selection when all combinations of phenotypes can evolve. Consequently, there is no theory that allows the diversity of observed pre-copulatory and post-copulatory inbreeding avoidance to be interpreted, and no clear hypotheses that could be tested through future empirical studies of individual systems or subsequent comparative analyses.

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundance) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Evolution of inbreeding avoidance can be driven by selection on both pre-copulatory and post-copulatory mechanisms, so simultaneous evolution of each phenotype might be defined by degeneracy, in addition to any direct costs on phenotypes. We used individual-based modelling to explore evolution of pre-copulatory and post-copulatory inbreeding strategy and polyandry given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given the pre-existence of the other. We thereby highlight fundamental but theoretically under-developed relationships between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding strategy. We present our model and simulation results as a tool for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.

>>>>

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "1 May 2017"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Inbreeding depression is widely hypothesised to drive adaptive evolution of pre-copulatory and post-copulatory mechanisms of inbreeding avoidance, which in turn are hypothesised to affect evolution of polyandry (i.e., female multiple mating). However, surprisingly little theory or modelling critically examines selection for pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, given evolutionary constraints and direct costs, or examines how evolution of inbreeding avoidance strategies might feed back to affect evolution of polyandry. Selection for post-copulatory inbreeding avoidance, but not for pre-copulatory inbreeding avoidance, requires polyandry, while interactions between pre-copulatory and post-copulatory inbreeding avoidance might cause functional redundancy (i.e., 'degeneracy') potentially generating complex evolutionary dynamics among inbreeding strategies and polyandry. We use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression and different evolutionary constraints and direct costs. We find that evolution of post-copulatory inbreeding avoidance increases selection for initially rare polyandry, and that selection for a costly inbreeding avoidance strategy was weak or negligible given a cost free alternative strategy. Further, fixed pre-copulatory inbreeding avoidance often completely precluded evolution of polyandry and therefore post-copulatory inbreeding avoidance, but fixed post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance. Evolution of inbreeding avoidance phenotypes and associated polyandry are likely affected by evolutionary feedbacks and degeneracy. All else being equal, while evolution of pre-copulatory inbreeding avoidance and associated low polyandry is more likely when pre-copulatory inbreeding avoidance is precluded or costly, evolution of post-copulatory inbreeding avoidance greatly facilitates evolution of costly polyandry.

@@

<!---

I'm confused by Ryan's point:

''It is 'required' in this sense but could a reviewer flag up that other forms of post-cop inbreeding avoidance could evolve without the need of polyandry?

I'm thinking in terms of sperm rejection/embryo inviability or something along those lines (such like you mention in the discussion), since females could then mate with a different male for subsequent events.  Very pedantic I know, but maybe a word or two of clarity re: polyandry is required for this specific type of post-cop inbreeding avoidance could cut this potential criticism off right away.''

If females mate with a different male, then it must be during a different bout of reproduction (else it's not polyandry as we define it). Perhaps what is meant is that inviability in one event effectively kills any parental investment that could affect future reproductive bouts? To me, it isn't really a different 'bout' if there is lingering parental investment that could affect current reproduction, but I guess I can see how a very pedantic reviewer could get confused. To make it completely air-tight (though a bit dense), I've added ''through fertilisation biased toward non-relatives''.

From the perspective of a breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), or by biasing fertilisation towards unrelated males after mating has occurred (i.e., post-copulatory inbreeding avoidance). Evolution of post-copulatory inbreeding avoidance through fertilisation biased toward non-relatives requires that females express some degree of polyandry, defined as mating with multiple males during a single reproductive bout. Such polyandry might initially be under selection and evolve if it allows for pre-copulatory inbreeding avoidance when females have the opportunity to mate with additional unrelated males following initial mating with a relative, assuming sufficiently strong inbreeding depression [@Duthie]. Polyandry can thereby facilitate pre-copulatory inbreeding avoidance even without any post-copulatory female choice or otherwise biased fertilisation among sperm provided by females' mates (i.e., under conditions of a 'fair raffle'). Overall, polyandry can therefore simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. Selection on polyandry to increase offspring fitness could explain its evolution in cases where multiple mating does not increase female reproductive success, or even reduces reproductive success and therefore imposes direct costs on polyandrous females. However, despite such widely-invoked hypotheses regarding the role of inbreeding depression on evolving strategies for pre-copulatory and post-copulatory inbreeding avoidance and hence polyandry, there is surprisingly little theory or modelling that critically examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or that examines how the evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry.

Despite the paucity of theory, numerous empirical studies on diverse species have tested for, and in some cases found evidence of, female inbreeding avoidance in systems where polyandry also occurs [@Tregenza2002; @Varian-Ramos2012; @Kingma2013; @Arct2015; but see @Reid2015a]. However, relatively few studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Pre-copulatory inbreeding avoidance has been reported in diverse species, spanning sweet potato weevils [*Cylas formicarius*, @Kuriwada2011], purple-crowned fairy-wrens [*Malurus coronatus*, @Kuriwada2011], and squinting bush brown butterflies [*Bicyclus anynana*, @Fischer2015], while post-copulatory inbreeding avoidance has been reported in red junglefowl [*Gallus gallus*, @Pizzari2004] and crickets [*Teleogryllus oceanicus*, @Simmons2006; *Gryllus bimaculatus*, @Bretman2009]. Evidence for both pre-copulatory and post-copulatory inbreeding avoidance is available across different studies of Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. Meanwhile, @Liu2014 found evidence of pre-copulatory but not post-copulatory inbreeding avoidance within a single study of cabbage beetles (*Colaphellus bowringi*). However, @Ala-Honkola2011 and @Tan2012 found no evidence for pre-copulatory or post-copulatory inbreeding avoidance in fruit flies (*Drosophila melanogaster*), respectively, and Reid et al. [-@Reid2014; -@Reid2015] found no net inbreeding avoidance in song sparrows (*Melospiza melodia*) despite strong inbreeding depression, suggesting neither pre-copulatory or post-copulatory inbreeding avoidance. Taken together, these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no theory that predicts what combinations of pre-copulatory and post-copulatory inbreeding avoidance and and associated polyandry should be favoured by selection when all combinations of phenotypes can evolve. Consequently, there is no theory that allows the diversity of observed pre-copulatory and post-copulatory inbreeding avoidance to be interpreted, and no clear hypotheses that could be tested through future empirical studies of individual systems or subsequent comparative analyses.

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundancy) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Evolution of inbreeding avoidance can be driven by selection on both pre-copulatory and post-copulatory mechanisms, so simultaneous evolution of each phenotype might be defined by degeneracy, in addition to any direct costs on phenotypes. We used individual-based modelling to explore evolution of pre-copulatory and post-copulatory inbreeding strategy and polyandry given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given the pre-existence of the other. We thereby highlight fundamental but theoretically under-developed relationships between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding strategy. We present our model and simulation results as a tool for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.
